Master Thesis

### 2013-05-22 my-work-rewritten.tex
Almost finished writing about defining signatures

>>>>

\subsubsection{Defining signatures}

In~\ref{sssec:sig}, we formally presented signatures as triples
$\mathopen{<}A, C, \tau\mathclose{>}$. The set $C$ can be easily
reconstructed as the domain of $\tau$ and the set $A$ is just the set of
all atomic types occurring in the range of $\tau$. Therefore, to define
the signature $\mathopen{<}A, C, \tau\mathclose{>}$, it is enough to
provide the function $\tau$, which in its set-theoretical realization is
just a set of pairs.  In relational programming, a set can be easily
encoded as a unary relation that answers the membership test. Therefore,
our goal will be to construct to answer this relation and this relation
will be our final representation of a signature.

We have our lexical database, a set of pairs of wordforms and hypertags,
and we want to arrive at a set of pairs of constants and their
types. This mapping decomposed into smaller parts by considering the
contribution of every part of the lexical database individually. What
will be the right model for the link between some part of the lexical
database and some part of our signature?

\begin{description}
  \item[Function of wordform] We could imagine that every different
    wordform of the lexical database will end up being mapped to some
    constant in the signature. However, this is unsuitable since the
    lexical database is a relation that can assign more than one
    distinct hypertag to a single wordform (e.g. with different part of
    speech) and we would like to have constants of different types for
    each of these hypertags.
  \item[Function of hypertag] Conversely, we might imagine that the
    hypertags of the lexical database are mapped via a function to the
    constants of the signature. Here we run into trouble when we
    consider variations in spelling or phonology. We might have two
    items in our lexical database that share the same hypertag but whose
    wordforms still differ. We would like to have two distinct constants
    in our signature then, so that when these two constants are mapped
    by the surface lexicon, they will yield two different strings.
  \item[Function of lexical entry] Since neither the wordform nor the
    hypertag is enough, we could consider the entire lexical entry (an
    element of the lexical database, a pair of a wordform and one of its
    hypertags). This solves both of the above problems but it is still
    not perfect. Consider the case of a determiner and some semantic
    signature. Since our link between elements of the lexical database
    and of our signature is a function, there must be a constant in the
    signature for every item of our lexical database. If you recall
    in~\ref{sssec:acg-sem}, the constants of the semantic signature
    $\Sigma_{Sem}$ contained predicates for the common nouns $man$ and
    $woman$ and for the verb $loves$ and some common logical
    furniture. However, it did not contain any constants which would
    correspond to the determiners $every$ and $some$.
  \item[Partial function of lexical entry] We could fix the issue above
    by using a partial function. This accounts for cases when a lexical
    entry has no constants to represent it in a signature. Nevertheless,
    this only fixes half of the problem. We will consider the single
    lexical entry which represents e.g. the 3rd person singular present
    tense form $loves$ of the transitive verb $love$. A legitimate
    account of subject/object scope ambiguity would be to consider two
    constants in our abstract syntactic signature that have distinct
    images in some semantic lexicon, one giving the outer scope to the
    subject and the other to the object. This means that in order to
    define such a signature, we would need to link a single lexical
    entry in our database to more than one constant in the signature.
  \item[Relation between lexical entry and constant] Finally,
    considering all the above common scenarios, the only mathematical
    model general enough to handle all of the above cases is that of a
    relation. In our system, this will correspond exactly to a relation
    that will be provided by the grammar author.
\end{description}

We have established that lexicons will be defined by a relation, a
relation between a lexical entry (a wordform and a hypertag) and a
constant (an identity\footnote{The identity of a constant is simply what
  distinguishes it from other constants of the same signature having the
  same type.} and a type). The identity of a lexical constant also has a
complex structure. We know that the lexical constants we define using
the relation described above are always linked to a lexical entry. This
lexical entry at least partially identifies the constant. For cases
where we have more than one constant per lexical entry (such as the
example of the two different scope readings of $loves$), we need some
extra piece of information to specify which constant we are referring to
(we call this extra piece of information a \emph{specifier}). This
specifier can have any shape that the grammar author desires. We can see
an example of a lexical constant in Figure~\ref{fig:lex-const}.

\begin{figure}
  % TODO: Center this.
  \centering
  \begin{verbatim}
{:type [-> NP [-> NP S]]
 :id {:lex-entry {:wordform "loves"
                  :hypertag ...}
      :spec {:scope :subject-wide}}}
  \end{verbatim}
  \caption{\label{fig:lex-const} An example of a lexical constant from a
    hypothetical syntactic signature (the hypertag is elided for brevity).}
\end{figure}

The system takes care of wiring up the relations and assembling the
signature. The grammar author's duty is to provide a 4-ary relation that
links the four parts of a lexical constant: the wordform, the hypertag,
the specifier and the type. This relation can be thought of as a
multi-valued ``function'' that receives as input a wordform and a
hypertag (that is, a single lexical entry) and outputs (possibly
multiple) pairs of specifiers and types. Since we are working in a
functional programming language that supports manipulation of
higher-order functions, we can provide helper functions which construct
these 4-ary relations for the grammar author in some simple cases. Our
library thus provides functions like \texttt{unitypedr}, which expects a
type and returns the 4-ary relation that assigns to every lexical item a
single constant whose specifier is $nil$ and whose type is the supplied
type, or \texttt{ht->typer}, which expects a mapping from hypertag
patterns to types and returns the 4-ary relation which tries to map the
hypertag against the patterns and yields constants with the
corresponding types and $nil$ specifiers. When other patterns emerge,
the grammar author is completely free to implement her new abstractions
using the full power of the functional programming language. See
Figure~\ref{fig:lex-sig-impl} for an example of implementing a
lexicalized version of the $\Sigma_{Synt}$ signature
from~\ref{sssec:example-sig} using the \texttt{ht->typer} abstraction.

\begin{figure}
  \centering
  \begin{verbatim}
(ht->typer {{:head {:cat "n"}}      'N
            {:head {:cat "v"
                    :trans "true"}} (-> 'NP 'NP 'S)
            {:head {:cat "det"}}    (-> 'N (-> 'NP 'S) 'S)})
  \end{verbatim}
  \caption{\label{fig:lex-sig-impl} A lexicalized version of the
    $\Sigma$%_{Synt} TODO TODO TODO
 signature from~\ref{sssec:example-sig}.}
\end{figure}

In the above paragraphs, we have been referring to the constants that we
were defining as \emph{lexical constants}. That was to distinguish them
from the \emph{non-lexical constants} that we will introduce now. The
reason for having non-lexical constants is that not every constant of
every signature is linked to (generated by) some item of our lexical
database. Consider, for example, the empty string $\epsilon$ and the
string concatenation operator $+$ of the $\Sigma_{String}$ signature we
gave in~\ref{sssec:example-sig} or the quantifiers and logical
connectives of the $\Sigma_{Sem}$ signature in~\ref{sssec:acg-sem}.

\subsection{Testing signatures and lexicons}
